---
title: "Spillover correction - comparison denoised non-denoised"
author: "Jacopo Umberto Verga"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
invisible(suppressMessages(c(library(imcRtools),
                 library(tidyverse),
                 library(patchwork),
                 library(scater),
                 library(BiocParallel),
                 library(CATALYST),
                 library(pheatmap),
                 library(cytomapper))))

multicore=MulticoreParam(workers=16)
```


# Import spatial dataset

I run the steinbock pipeline with the original data. Now I am going to run the spillover correction, observe the results and compare with the denoised data

```{r load_spe}
spe_dn <- readRDS("/home/rstudio/workspace/data/mydata/spe.rds")
spe_og <- readRDS("/home/rstudio/workspace/data/non_denoised/spe.rds")
```




## Single-cell data compensation

The `CATALYST` package can be used to perform spillover compensation on the
**single-cell mean intensities**. Here, the `SpatialExperiment` object generated
in in the previous Session is read in. The `CATALYST` package requires an entry
to `rowData(spe)$channel_name` for the `compCytof` function to run. This entry
should contain the metal isotopes in the form (mt)(mass)Di (e.g., `Sm152Di` for
Samarium isotope with the atomic mass 152). 

The `compCytof` function performs channel spillover compensation on the mean
pixel intensities per channel and cell. Here, we will not overwrite the assays
in the `SpatialExperiment` object to later highlight the effect of compensation.
As shown in the previous session, also the compensated counts are
asinh-transformed using a cofactor of 1.

```{r single-cell-compensation}

sm <- read.csv('/home/rstudio/workspace/data/compensation_matrix_hyperion/compensation_matrix.csv', header = TRUE,  check.names = FALSE, row.names = '')
sm <- sm[rownames(sm)]
sm[is.na(sm)] <- 0
sm <- sm/100
sm <- as.matrix(sm)
diag(sm) <- 1


spillover_correction <- function(spe, sm){
    channel_names <- paste0(gsub("[()]","", rowData(spe)$channel),"Di")
    rowData(spe)$channel_name <- channel_names
    rownames(sm) <- channel_names
    colnames(sm) <- channel_names
    
    isotope_list <- CATALYST::isotope_list
    isotope_list$Ar <- 80
    spe <- compCytof(spe, sm, 
                     transform = TRUE, cofactor = 1,
                     isotope_list = isotope_list, 
                     overwrite = FALSE)
    return(spe)
}

```

```{r}

spe_og <- spillover_correction(spe_og, sm)
spe_dn <- spillover_correction(spe_dn, sm)

```     



To check the effect of channel spillover compensation, the expression of markers
that are affected by spillover (e.g., GZMK in channel Gd(158) and aSMA in
channel Nd(142)) can be visualized in form of scatter plots before and after
compensation.

```{r visualize-single-cell-spillover, message=FALSE}
library(dittoSeq)
library(patchwork)

spill1='aSMA'
spill2='Granzyme_K'

corr = cor(as.data.frame(t(exprs(spe_og)))[c(spill1, spill2)])[2,1]
before_og <- dittoScatterPlot(spe_og, x.var = spill1, y.var = spill2,
                           assay.x = "exprs", assay.y = "exprs") +
    ggtitle(paste0("Before compensation - Original - ", round(corr,2)))

mat = as.data.frame(t(spe_og@assays@data$compexprs))
meta = rowData(spe_og)['channel_name']
colnames(mat) <-rownames(meta)[match(colnames(mat), meta$channel_name)]
corr = cor(mat[c(spill1, spill2)])[2,1]
after_og <- dittoScatterPlot(spe_og, x.var = spill1, y.var = spill2,
                          assay.x = "compexprs", assay.y = "compexprs") +
    ggtitle(paste0("After compensation - Original - ", round(corr,2)))


corr = cor(as.data.frame(t(exprs(spe_dn)))[c(spill1, spill2)])[2,1]
before_dn <- dittoScatterPlot(spe_dn, x.var = spill1, y.var = spill2,
                           assay.x = "exprs", assay.y = "exprs") +
    ggtitle(paste0("Before compensation - Denoised - ", round(corr,2)))


mat = as.data.frame(t(spe_dn@assays@data$compexprs))
meta = rowData(spe_og)['channel_name']
colnames(mat) <-rownames(meta)[match(colnames(mat), meta$channel_name)]
corr = cor(mat[c(spill1, spill2)])[2,1]
after_dn <- dittoScatterPlot(spe_dn, x.var = spill1, y.var = spill2,
                          assay.x = "compexprs", assay.y = "compexprs") +
    ggtitle(paste0("After compensation - Denoised - ", round(corr,2)))
(before_og + after_og)/(before_dn + after_dn)

```






```{r fig.height=12, fig.width=15}
library(reshape2)
 

mat = t(exprs(spe_og))
corr_mat <- cor(mat)
dist <- as.dist((1-corr_mat)/2)
hc <- hclust(dist)
clusterCut <- cutree(hc,3)
# Convert named integer vector to a list of observations per cluster
cluster_list_og_pre <- split(names(clusterCut), clusterCut)


corr_mat <-corr_mat[hc$order, hc$order]
melted_corr_mat <- melt(corr_mat)
p1 = ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2, fill=value)) + 
geom_tile() + 
    labs(title='Correlation before compensation - Original')+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


mat = t(spe_og@assays@data$compexprs)
meta = rowData(spe_og)['channel_name']
colnames(mat) <-rownames(meta)[match(colnames(mat), meta$channel_name)]
corr_mat <- cor(mat)
dist <- as.dist((1-corr_mat)/2)
 hc <- hclust(dist)
 
clusterCut <- cutree(hc,3)
# Convert named integer vector to a list of observations per cluster
cluster_list_og_post <- split(names(clusterCut), clusterCut)
corr_mat <-corr_mat[hc$order, hc$order]
melted_corr_mat <- melt(corr_mat)
p2 = ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2, fill=value)) + 
geom_tile() + 
    labs(title='Correlation after compensation - Original')+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



mat = t(exprs(spe_dn))
corr_mat <- cor(mat)
dist <- as.dist((1-corr_mat)/2)
 hc <- hclust(dist)
 
clusterCut <- cutree(hc,3)
# Convert named integer vector to a list of observations per cluster
cluster_list_dn_pre <- split(names(clusterCut), clusterCut)
corr_mat <-corr_mat[hc$order, hc$order]
melted_corr_mat <- melt(corr_mat)
p3 = ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2, fill=value)) + 
geom_tile() + 
    labs(title='Correlation before compensation - Denoised')+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


mat = t(spe_dn@assays@data$compexprs)
meta = rowData(spe_og)['channel_name']
colnames(mat) <-rownames(meta)[match(colnames(mat), meta$channel_name)]
corr_mat <- cor(mat)
dist <- as.dist((1-corr_mat)/2)
hc <- hclust(dist)
clusterCut <- cutree(hc,3)
# Convert named integer vector to a list of observations per cluster
cluster_list_dn_post <- split(names(clusterCut), clusterCut)
corr_mat <-corr_mat[hc$order, hc$order]
melted_corr_mat <- melt(corr_mat)
p4 = ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2, fill=value)) + 
geom_tile() + 
    labs(title='Correlation after compensation - Denoised')+
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


(p1+p2)/(p3+p4)
```


```{r}
# Function to transform vector into comma-separated string
vector_to_string <- function(vec) {
  return(paste(vec, collapse = ", "))
}

# Combine lists into a named list of lists
named_lists <- list(list1 =  lapply(cluster_list_og_pre, vector_to_string), 
                    list2 =  lapply(cluster_list_og_post, vector_to_string), 
                    list3 =  lapply(cluster_list_dn_pre, vector_to_string), 
                    list4 =  lapply(cluster_list_dn_post, vector_to_string))

# Extract values for common names (1, 2, 3, 4) and store in a matrix
num_lists <- length(named_lists)
common_names <- c("1", "2", "3")

# Initialize a matrix to store the values
data_matrix <- matrix("", nrow = num_lists, ncol = length(common_names))

# Loop over each list and extract values for common names
for (i in 1:num_lists) {
  list_elements <- unlist(named_lists[[i]])  # Flatten the list elements
  values <- sapply(common_names, function(name) list_elements[[name]])  # Extract values for common names
  data_matrix[i, ] <- values
}




# Convert the matrix to a DataFrame with appropriate column names
df <- as.data.frame(data_matrix)
colnames(df) <- common_names

# Print the resulting DataFrame
df
```


```{r save-objects-read-data}
saveRDS(spe_og, "/home/rstudio/workspace/data/non_denoised/spe_spillover_corrected.rds")
saveRDS(spe_dn, "/home/rstudio/workspace/data/mydata/spe_spillover_corrected.rds")
```

## Image compensation

The [cytomapper](https://github.com/BodenmillerGroup/cytomapper) package allows channel
spillover compensation directly on **multi-channel images**. 
The `compImage` function takes a `CytoImageList` object and the estimated
spillover matrix as input. 

At this point, we can read in the `CytoImageList` object containing multi-channel
images as generated in the previous session.
The `channelNames` need to be set according to their metal isotope in the form
(mt)(mass)Di and therefore match `colnames(sm)`.

```{r read-in-image, eval=FALSE, message=FALSE, include=FALSE}
library(cytomapper)

images <- readRDS("/home/rstudio/workspace/data/non_denoised/images.rds")
masks <- readRDS("/home/rstudio/workspace/data/non_denoised/masks.rds")
channelNames(images) <- rowData(spe)$channel_name
```

The CATALYST package provides the `adaptSpillmat` function that corrects the
spillover matrix in a way that rows and columns match a predefined set of
metals. Please refer to `?compCytof` for more information how metals in the
spillover matrix are matched to acquired channels in the `SingleCellExperiment`
object.

The spillover matrix can now be adapted to exclude channels that are not part of
the measurement (`keep == 0`).

```{r eval=FALSE, include=FALSE}
library(tiff)
panel <- read.csv("/home/rstudio/workspace/data/non_denoised//panel.csv")

channel_names <- paste0(gsub("[()]","", rowData(spe)$channel),"Di")
rowData(spe)$channel_name <- channel_names
rownames(sm) <- channel_names
colnames(sm) <- channel_names

isotope_list <- CATALYST::isotope_list
isotope_list$Ar <- 80

adapted_sm <- adaptSpillmat(sm, paste0(gsub("[()]","", panel$channel[panel$keep == 1]),"Di"), 
                            isotope_list = isotope_list)
```

The adpated spillover matrix now matches the `channelNames` of the
`CytoImageList` object and can be used to perform pixel-level spillover
compensation.

```{r image-compensation, eval=FALSE, message=FALSE, include=FALSE}
library(BiocParallel)
library(parallel)

images_comp <- compImage(images, adapted_sm, 
                         BPPARAM = MulticoreParam(workers = 16))
```

As a sanity check, we will visualize the image before and after compensation:

```{r image-visualization, eval=FALSE, message=FALSE, include=FALSE}
# Before compensation
plotPixels(images[5], colour_by = "Nd142Di", 
           image_title = list(text = "Nd142Di (aSMA) - before", position = "topleft"), 
           legend = NULL, bcg = list(Nd142Di = c(0, 4, 1)))
plotPixels(images[5], colour_by = "Gd158Di", 
           image_title = list(text = "Gd158Di (GZMK) - before", position = "topleft"), 
           legend = NULL, bcg = list(Gd158Di = c(0, 4, 1)))

# After compensation
plotPixels(images_comp[5], colour_by = "Nd142Di",
           image_title = list(text = "Nd142Di (aSMA) - after", position = "topleft"), 
           legend = NULL, bcg = list(Nd142Di = c(0, 4, 1)))
plotPixels(images_comp[5], colour_by = "Gd158Di", 
           image_title = list(text = "Gd158Di (GZMK) - after", position = "topleft"),
           legend = NULL, bcg = list(Gd158Di = c(0, 4, 1)))
```

For convenience, we will re-set the `channelNames` to their biological targtes:

```{r re-set-channels, eval=FALSE, message=FALSE, include=FALSE}
channelNames(images_comp) <- rownames(spe)
```



```{r save-images-read-data, eval=FALSE, message=FALSE, include=FALSE}
saveRDS(images_comp, "/home/rstudio/workspace/data/non_denoised/images_comp.rds")

```