---
title: "Batch_correction - OG data"
author: "Jacopo Umberto verga"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Single Cell object

Considering  the lower correlation between channels with spillover in the previous notebook, I am going to use the spillover corrected original images for the next analysis



```{r spatial-load-data, message=FALSE, cache=FALSE}
library(SpatialExperiment)
library(tidyverse)
library(cytomapper)
library(imcRtools)
library(stringr)
library(scater)
library(BiocParallel)
multicore=MulticoreParam(workers=16)

spe <- readRDS("../data/non_denoised/spe_spillover_corrected.rds")

# Overwrite assays woth the corrected ones
assay(spe, "counts") <- assay(spe, "compcounts") 
assay(spe, "exprs") <- assay(spe, "compexprs") 
assay(spe, "compcounts") <- assay(spe, "compexprs") <- NULL
```

To run dimensionality reduction and clustering I am going to use:
- A subset of cells (10%) to reduce the computational times
-  subset of markers to reduce the noise


```{r select-features, , cache=FALSE}

meta_ds <- read.csv("../Session4_single_cell_analysis/adata_ds_meta.csv")

# spe_bkp <- spe
# spe <- spe[,colnames(spe) %in% meta_ds$X]



features_oi <- c('CD45','CD45RO','CD3','CD4','CD8','CD20','CD27','CD138','CD14','CD16','CD56','CD68','CD11c','CD11b','CD31','aSMA')

rowData(spe)$use_channel <- rownames(spe) %in% features_oi
rowData(spe)['use_channel']
```

# Dimensionality reduction and visualization

Before doing the batch correction, I will try to run PCA, KNN, UMAP of non-integrated data:

```{r dimred, message=FALSE, cache=FALSE}
set.seed(12345)
spe <- runUMAP(spe, subset_row = rowData(spe)$use_channel, exprs_values = "exprs", BPPARAM=multicore) 
```

After dimensionality reduction, the low-dimensional embeddings are stored in the 
`reducedDim` slot.

```{r show-dimred-slot, cache=FALSE}
reducedDims(spe)
head(reducedDim(spe, "UMAP"))
```



```{r visualizing-dimred-1, fig.height=12, fig.width=8, message=FALSE, paged.print=FALSE, cache=FALSE}
library(patchwork)
library(dittoSeq)
library(viridis)

# visualize patient id 
p1 <- dittoDimPlot(spe, var = "sample_id", reduction.use = "UMAP", size = 0.2, legend.show = FALSE) + 
    ggtitle("Patient ID on UMAP")
p2 <- dittoDimPlot(spe, var = "patient_n", reduction.use = "UMAP", size = 0.2, legend.size = 3) + 
    ggtitle("Slide ID on UMAP")

# visualize indication
p3 <- dittoDimPlot(spe, var = "timepoint", reduction.use = "UMAP", size = 0.2, legend.size = 3) + 
    scale_color_manual(values = metadata(spe)$color_vectors$timepoint) +
    ggtitle("Timepoint on UMAP")
p4 <- dittoDimPlot(spe, var = "CD3", reduction.use = "UMAP", size = 0.2, legend.size = 3) + 
    ggtitle("CD3 on UMAP")
# visualize indication
p5 <- dittoDimPlot(spe, var = "CD138", reduction.use = "UMAP", size = 0.2, legend.size = 3) + 
    ggtitle("CD138 on UMAP")
p6 <- dittoDimPlot(spe, var = "CD31", reduction.use = "UMAP", size = 0.2, legend.size = 3) + 
    ggtitle("CD31 on UMAP")

(p1 + p2) / (p3 + p4) / (p5+p6)
```


We can see a lil batch effect and not good division between features. Batch correction on slide ID.


# Batch correction


## fastMNN correction


```{r batch-correction-fastMNN, message=FALSE, warning=FALSE, cache=FALSE}
library(batchelor)
set.seed(12345)
out <- fastMNN(spe, batch = spe$sample_id,
               auto.merge = TRUE,
               subset.row = rowData(spe)$use_channel,
               assay.type = "exprs",
               BPPARAM = multicore)

# Transfer the correction results to the main spe object
reducedDim(spe, "fastMNN") <- reducedDim(out, "corrected")
```



```{r batch-correction-fastMNN-QC, message=FALSE, cache=FALSE}
merge_info <- metadata(out)$merge.info 

DataFrame(left = merge_info$left,
          right = merge_info$right,
          batch.size = merge_info$batch.size,
          max_lost_var = rowMax(merge_info$lost.var))
```




### Visualization


```{r dimred-batch-correction-fastMNN, message=FALSE, warning=FALSE, cache=FALSE}
library(scater)

set.seed(12345)
spe <- runUMAP(spe, dimred= "fastMNN", name = "UMAP_mnnCorrected") 
```


```{r visualizing-dimred-2, fig.height=12, fig.width=8, message=FALSE, paged.print=FALSE, cache=FALSE}
library(patchwork)
library(dittoSeq)
library(viridis)

# visualize patient id 
p1 <- dittoDimPlot(spe, var = "sample_id", reduction.use = "UMAP_mnnCorrected", size = 0.2, legend.show = FALSE) + 
    ggtitle("Patient ID on UMAP")
p2 <- dittoDimPlot(spe, var = "patient_n", reduction.use = "UMAP_mnnCorrected", size = 0.2, legend.size = 3) + 
    ggtitle("Slide ID on UMAP")

# visualize indication
p3 <- dittoDimPlot(spe, var = "timepoint", reduction.use = "UMAP_mnnCorrected", size = 0.2, legend.size = 3) + 
    scale_color_manual(values = metadata(spe)$color_vectors$timepoint) +
    ggtitle("Timepoint on UMAP")
p4 <- dittoDimPlot(spe, var = "CD3", reduction.use = "UMAP_mnnCorrected", size = 0.2, legend.size = 3) + 
    ggtitle("CD3 on UMAP")
# visualize indication
p5 <- dittoDimPlot(spe, var = "CD138", reduction.use = "UMAP_mnnCorrected", size = 0.2, legend.size = 3) + 
    ggtitle("CD138 on UMAP")
p6 <- dittoDimPlot(spe, var = "CD31", reduction.use = "UMAP_mnnCorrected", size = 0.2, legend.size = 3) + 
    ggtitle("CD31 on UMAP")

(p1 + p2) / (p3 + p4) / (p5+p6)
```



```{r visualizing-batch-correction-fastMNN-2, warning=FALSE, message=FALSE, fig.height=8, cache=FALSE}
library(cowplot)
# Before correction
plot_list <- multi_dittoDimPlot(spe, var = features_oi, reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2, list.out = TRUE) 
plot_list <- lapply(plot_list, function(x) x + scale_color_viridis())
plot_grid(plotlist = plot_list) 

# After correction
plot_list <- multi_dittoDimPlot(spe, var = features_oi, reduction.use = "UMAP_mnnCorrected", 
                   assay = "exprs", size = 0.2, list.out = TRUE) 
plot_list <- lapply(plot_list, function(x) x + scale_color_viridis())
plot_grid(plotlist = plot_list) 
```



## Harmony

```{r run_harmony, cache=FALSE}
library(harmony)
library(BiocSingular)

spe <- runPCA(spe, 
              subset_row = rowData(spe)$use_channel, 
              exprs_values = "exprs", 
              ncomponents = 15,
              BSPARAM = ExactParam())

set.seed(12345)
out <- RunHarmony(spe, group.by.vars = "sample_id",
                  lambda=.1,
                  plot_convergence=TRUE)

# Check that order of cells is the same
stopifnot(all.equal(colnames(spe), colnames(out)))

reducedDim(spe, "harmony") <- reducedDim(out, "HARMONY")
```


### Visualization


```{r dimred-batch-correction-harmony, message=FALSE, warning=FALSE, cache=FALSE}
library(scater)

set.seed(12345)
spe <- runUMAP(spe, dimred= "harmony", name = "UMAP_harmony") 
```

Next, we visualize the corrected UMAP while overlaying patient IDs.

```{r visualizing-dimred-3, fig.height=12, fig.width=8, message=FALSE, paged.print=FALSE, cache=FALSE}
library(patchwork)
library(dittoSeq)
library(viridis)

# visualize patient id 
p1 <- dittoDimPlot(spe, var = "sample_id", reduction.use = "UMAP_harmony", size = 0.2, legend.show = FALSE) + 
    ggtitle("Patient ID on UMAP")
p2 <- dittoDimPlot(spe, var = "patient_n", reduction.use = "UMAP_harmony", size = 0.2, legend.size = 3) + 
    ggtitle("Slide ID on UMAP")

# visualize indication
p3 <- dittoDimPlot(spe, var = "timepoint", reduction.use = "UMAP_harmony", size = 0.2, legend.size = 3) + 
    scale_color_manual(values = metadata(spe)$color_vectors$timepoint) +
    ggtitle("Timepoint on UMAP")
p4 <- dittoDimPlot(spe, var = "CD3", reduction.use = "UMAP_harmony", size = 0.2, legend.size = 3) + 
    ggtitle("CD3 on UMAP")
# visualize indication
p5 <- dittoDimPlot(spe, var = "CD138", reduction.use = "UMAP_harmony", size = 0.2, legend.size = 3) + 
    ggtitle("CD138 on UMAP")
p6 <- dittoDimPlot(spe, var = "CD31", reduction.use = "UMAP_harmony", size = 0.2, legend.size = 3) + 
    ggtitle("CD31 on UMAP")

(p1 + p2) / (p3 + p4) / (p5+p6)
```



```{r visualizing-batch-correction-harmony-2, warning=FALSE, message=FALSE, fig.height=8, cache=FALSE}

# Before correction
plot_list <- multi_dittoDimPlot(spe, var = features_oi, reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2, list.out = TRUE) 
plot_list <- lapply(plot_list, function(x) x + scale_color_viridis())
plot_grid(plotlist = plot_list) 

# After correction
plot_list <- multi_dittoDimPlot(spe, var = features_oi, reduction.use = "UMAP_harmony", 
                   assay = "exprs", size = 0.2, list.out = TRUE) 
plot_list <- lapply(plot_list, function(x) x + scale_color_viridis())
plot_grid(plotlist = plot_list) 
```

# Save object

```{r save-objects}
saveRDS(spe, "../data/non_denoised/spe_batch_corrected_sample_id.Rds")
```

## Session info

```{r}
sessionInfo()
```
