---
title: 'Annotating clusters'
author: "Jacopo Umberto verga"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
    
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Single Cell object

I assigned the major cell types using a XGBoost classifier trained with a subset of cells manually annotated. Now I am going to load the labels here and assign sub-types with Gaussian Mixture Models.



```{r spatial-load-data, message=FALSE, cache=FALSE}
library(SpatialExperiment)
library(SingleCellExperiment)
library(tidyverse)
library(cytomapper)
library(imcRtools)
library(stringr)
library(scater)
library(BiocParallel)
library(viridis)
multicore=MulticoreParam(workers=16)


spe <- readRDS("../data/non_denoised/spe_batch_corrected_exprs.Rds")

cell_labels <- read.csv('../data/non_denoised/csv_tables_all_cells/major_labels.csv',row.names=1)
spe$major_celltype <- cell_labels$major_celltype
```


```{r fig.height=6, fig.width=15, cache=FALSE}
library(gridExtra)
library(grid)
library(pheatmap)

mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```

```{r}
props=as.data.frame(table(spe$major_celltype))%>%
    mutate(Freq=Freq/sum(Freq))
ggplot(props, aes(x=Var1, y=Freq, fill=Var1))+
    geom_bar(stat='identity')+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),legend.position = "none") +
    labs(x='Cell type',y='Proportion', title='Cell types proportions after XGBoost classification') 
```




We have quite mixed cell types, now I am going to refine the classification. First I start with Megakaryocytes, they have some expression of myeloid markers.

```{r}
update_label <- function(old_object, new_object){
    old_object[,colnames(old_object) %in% colnames(new_object)]$major_celltype <- new_object$mclust_label
    return(old_object)
    
}
    
```



# Cleaning up Megakaryocytes:

Many positive markers, Use CD31, CD68, CD14, CD16


 
```{r}
spe_megakaryo = spe[,spe$major_celltype == 'Megakaryocytes']
```
 
 
```{r}
library(mclust)
mat <- as.data.frame(t(exprs(spe_megakaryo))[,c('CD31','CD68','CD16','CD163')])
set.seed(12345)
mBIC <- mclust::mclustBIC(mat, G=2:10)
set.seed(12345)
clusters = Mclust(mat, x=mBIC)
spe_megakaryo$mclust_cluster = clusters$classification
heatmap_data = clusters$parameters$mean
colnames(heatmap_data) <- seq(1:dim(heatmap_data)[2])
pheatmap(heatmap_data, main="Not scaled", display_numbers=TRUE,fontsize_number=10,show_colnames=TRUE,fontsize_col=20,angle_col=0)
pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=20,angle_col=0)
```
 
 

 
```{r}
cluster_celltype <- recode(spe_megakaryo$mclust_cluster,
                           "1"='Megakaryocytes',
                           "2"='Other myeloid',
                           "3"='Other myeloid',
                           "4"='Other myeloid',
                           "5"='Other myeloid',
                           "6"='Other myeloid',
                           "7"='Megakaryocytes',
                           "8"='Other myeloid',
                           "9"='Other myeloid',
                           "10"='Megakaryocytes')

spe_megakaryo$mclust_label <-cluster_celltype

colnames(heatmap_data) <- recode(colnames(heatmap_data),
                           "1"='Megakaryocytes',
                           "2"='Other myeloid',
                           "3"='Other myeloid',
                           "4"='Other myeloid',
                           "5"='Other myeloid',
                           "6"='Other myeloid',
                           "7"='Megakaryocytes',
                           "8"='Other myeloid',
                           "9"='Other myeloid',
                           "10"='Megakaryocytes')

spe_megakaryo$mclust_label <-cluster_celltype

pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=10,angle_col=45)
```


```{r}


spe <- update_label(spe, spe_megakaryo)
```





```{r fig.height=6, fig.width=15, cache=FALSE}
mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```


# Cleaning up Other myeloid:



 
```{r}
spe_myelo_a = spe[,spe$major_celltype == 'Other myeloid']
```
 
 
```{r}
mat <- as.data.frame(t(exprs(spe_myelo_a))[,c('CD14','CD68','CD16', 'CD11b')])
set.seed(12345)
mBIC <- mclust::mclustBIC(mat, G=2:10)
set.seed(12345)
clusters = Mclust(mat, x=mBIC)
spe_myelo_a$mclust_cluster = clusters$classification
heatmap_data = clusters$parameters$mean
colnames(heatmap_data) <- seq(1:dim(heatmap_data)[2])
pheatmap(heatmap_data, main="Not scaled", display_numbers=TRUE,fontsize_number=10,show_colnames=TRUE,fontsize_col=20,angle_col=0)
pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=4,show_colnames=TRUE,fontsize_col=20,angle_col=0)
```
 
 

 
```{r}
cluster_celltype <- recode(spe_myelo_a$mclust_cluster,
                           "1"='Macro',
                           "2"='Other myeloid',
                           "3"='Other myeloid',
                           "4"='Other myeloid',
                           "5"='Other myeloid',
                           "6"='Other myeloid',
                           "7"='Other myeloid',
                           "8"='Other myeloid',
                           "9"='Monocytes',
                           "10"='Macro')

spe_myelo_a$mclust_label <-cluster_celltype

colnames(heatmap_data) <- recode(colnames(heatmap_data),
                           "1"='Macro',
                           "2"='Other myeloid',
                           "3"='Other myeloid',
                           "4"='Other myeloid',
                           "5"='Other myeloid',
                           "6"='Other myeloid',
                           "7"='Other myeloid',
                           "8"='Other myeloid',
                           "9"='Monocytes',
                           "10"='Macro')

spe_myelo_a$mclust_label <-cluster_celltype

pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=4,show_colnames=TRUE,fontsize_col=10,angle_col=45)
```


```{r}

spe <- update_label(spe, spe_myelo_a)
```




```{r fig.height=6, fig.width=15, cache=FALSE}
mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```


```{r}
table(spe$major_celltype)
```



## Define other myeloid:

CD117, CD11c, CD11b, HLA-DR

 
```{r}
spe_mast_a = spe[,spe$major_celltype == 'Other myeloid']
```
 
 
```{r}
mat <- as.data.frame(t(exprs(spe_mast_a))[,c('CD117', 'CD11c', 'CD11b', 'HLA-DR')])
set.seed(12345)
mBIC <- mclust::mclustBIC(mat, G=2:10)
set.seed(12345)
clusters = Mclust(mat, x=mBIC)
```

```{r}
spe_mast_a$mclust_cluster = clusters$classification
heatmap_data = clusters$parameters$mean
colnames(heatmap_data) <- seq(1:dim(heatmap_data)[2])
pheatmap(heatmap_data, main="Not scaled", display_numbers=TRUE,fontsize_number=10,show_colnames=TRUE,fontsize_col=20,angle_col=0)
pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=3,show_colnames=TRUE,fontsize_col=20,angle_col=0)
```
 


 
```{r}
cluster_celltype <- recode(spe_mast_a$mclust_cluster,
                           "1"='DC',
                           "2"='Other myeloid',
                           "3"='DC',
                           "4"='Other myeloid',
                           "5"='Other myeloid',
                           "6"='Other myeloid',
                           "7"='Other myeloid',
                           "8"='Other myeloid',
                           "9"='Other myeloid',
                           "10"='Other myeloid')

spe_mast_a$mclust_label <-cluster_celltype

colnames(heatmap_data) <- recode(colnames(heatmap_data),
                           "1"='DC',
                           "2"='Other myeloid',
                           "3"='DC',
                           "4"='Other myeloid',
                           "5"='Other myeloid',
                           "6"='Other myeloid',
                           "7"='Other myeloid',
                           "8"='Other myeloid',
                           "9"='Other myeloid',
                           "10"='Other myeloid')


spe_mast_a$mclust_label <-cluster_celltype

pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=3,show_colnames=TRUE,fontsize_col=10,angle_col=45)
```


```{r}
spe <- update_label(spe, spe_mast_a)
```




```{r fig.height=6, fig.width=15, cache=FALSE}
mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```

```{r}
table(spe$major_celltype)
```








## Macro:


 
```{r}
spe_mono_macro = spe[,spe$major_celltype == 'Macro']
```
 
 
```{r}
mat <- as.data.frame(t(exprs(spe_mono_macro))[,c('CD14', 'CD16', 'CD68', 'CD163','HLA-DR')])
set.seed(12345)
mBIC <- mclust::mclustBIC(mat, G=2:10)
set.seed(12345)
clusters = Mclust(mat, x=mBIC)
spe_mono_macro$mclust_cluster = clusters$classification
```

```{r}
heatmap_data = clusters$parameters$mean
colnames(heatmap_data) <- seq(1:dim(heatmap_data)[2])
pheatmap(heatmap_data, main="Not scaled", display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=20,angle_col=0)
pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=20,angle_col=0)
```
 
 

 
```{r}
cluster_celltype <- recode(spe_mono_macro$mclust_cluster,
                           "1"='Macrophages-hla',
                           "2"='Macrophages',
                           "3"='Macrophages',
                           "4"='Macrophages',
                           "5"='Macrophages-hla',
                           "6"='Macrophages',
                           "7"='Macrophages-CD163 neg',
                           "8"='Macrophages-hla',
                           "9"='Macrophages-hla',
                           "10"='Macrophages')

spe_mono_macro$mclust_label <-cluster_celltype

colnames(heatmap_data) <- recode(colnames(heatmap_data),
                           "1"='Macrophages-hla',
                           "2"='Macrophages',
                           "3"='Macrophages',
                           "4"='Macrophages',
                           "5"='Macrophages-hla',
                           "6"='Macrophages',
                           "7"='Macrophages-CD163 neg',
                           "8"='Macrophages-hla',
                           "9"='Macrophages-hla',
                           "10"='Macrophages')


spe_mono_macro$mclust_label <-cluster_celltype

pheatmap(heatmap_data, scale='none', main='Non scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=10,angle_col=45)
```


```{r}
spe <- update_label(spe, spe_mono_macro)
```




```{r fig.height=6, fig.width=15, cache=FALSE}
mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```



# Monocytes:


 
```{r}
spe_mono = spe[,spe$major_celltype == 'Monocytes']
```
 
 
```{r}
mat <- as.data.frame(t(exprs(spe_mono))[,c('CD14', 'CD16', 'CCR4', 'CCR6', 'HLA-DR','CD45','CD11c')])
set.seed(12345)
mBIC <- mclust::mclustBIC(mat, G=2:10)
set.seed(12345)
clusters = Mclust(mat, x=mBIC)
```

```{r}
spe_mono$mclust_cluster = clusters$classification
heatmap_data = clusters$parameters$mean
colnames(heatmap_data) <- seq(1:dim(heatmap_data)[2])
pheatmap(heatmap_data, main="Not scaled", display_numbers=TRUE,fontsize_number=10,cutree_cols=1,show_colnames=TRUE,fontsize_col=20,angle_col=0)
pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=1,show_colnames=TRUE,fontsize_col=20,angle_col=0)
```
 
 

 
```{r}
cluster_celltype <- recode(spe_mono$mclust_cluster,
                           "1"='Non Classical Monocytes',
                           "2"='Intermediate Monocytes',
                           "3"='Classical Monocytes',
                           "4"='Non Classical Monocytes',
                           "5"='Classical Monocytes',
                           "6"='Classical Monocytes',
                           "7"='Non Classical Monocytes',
                           "8"='Non Classical Monocytes',
                           "9"='Non Classical Monocytes',
                           "10"='Classical Monocytes')

spe_mono$mclust_label <-cluster_celltype

colnames(heatmap_data) <- recode(colnames(heatmap_data),
                           "1"='Non Classical Monocytes',
                           "2"='Intermediate Monocytes',
                           "3"='Classical Monocytes',
                           "4"='Non Classical Monocytes',
                           "5"='Classical Monocytes',
                           "6"='Classical Monocytes',
                           "7"='Non Classical Monocytes',
                           "8"='Non Classical Monocytes',
                           "9"='Non Classical Monocytes',
                           "10"='Classical Monocytes')


spe_mono$mclust_label <-cluster_celltype

pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=1,show_colnames=TRUE,fontsize_col=10,angle_col=45)
```


```{r}
spe <- update_label(spe, spe_mono)
```




```{r fig.height=6, fig.width=15, cache=FALSE}
mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```




# T:


 
```{r}
spe_T = spe[,spe$major_celltype == 'T']
```
 
 
```{r}
mat <- as.data.frame(t(exprs(spe_T))[,c('CD4', 'CD8')])
set.seed(12345)
mBIC <- mclust::mclustBIC(mat, G=2:10)
set.seed(12345)
clusters = Mclust(mat, x=mBIC)
spe_T$mclust_cluster = clusters$classification
heatmap_data = clusters$parameters$mean
colnames(heatmap_data) <- seq(1:dim(heatmap_data)[2])
pheatmap(heatmap_data, main="Not scaled", display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=20,angle_col=0)
pheatmap(heatmap_data, scale='row', main='Scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=20,angle_col=0)
```
 
 

 
```{r}
cluster_celltype <- recode(spe_T$mclust_cluster,
                           "1"='CD4 T',
                           "2"='CD4 T',
                           "3"='CD4 T',
                           "4"='CD4 T',
                           "5"='CD8 T',
                           "6"='CD4 T',
                           "7"='CD4 T',
                           "8"='CD8 T',
                           "9"='CD4 T')

spe_T$mclust_label <-cluster_celltype

colnames(heatmap_data) <- recode(colnames(heatmap_data),
                           "1"='CD4 T',
                           "2"='CD4 T',
                           "3"='CD4 T',
                           "4"='CD4 T',
                           "5"='CD8 T',
                           "6"='CD4 T',
                           "7"='CD4 T',
                           "8"='CD8 T',
                           "9"='CD4 T')


spe_T$mclust_label <-cluster_celltype

pheatmap(heatmap_data, scale='none', main='Non scaled', display_numbers=TRUE,fontsize_number=10,cutree_cols=2,show_colnames=TRUE,fontsize_col=10,angle_col=45)
```


```{r}
spe <- update_label(spe, spe_T)
```




```{r fig.height=6, fig.width=15, cache=FALSE}
mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```








```{r eval=FALSE, fig.height=6, fig.width=15, cache=FALSE, include=FALSE}
mat = subset(as.data.frame(t(exprs(spe))), select = -c(Histone_H3, ICSK1, ICSK2))
mat$cluster = spe$major_celltype
mat <- aggregate(. ~ cluster, data = mat, FUN = mean)
rownames(mat) <- mat$cluster
mat$cluster <- NULL
p1 = pheatmap(t(mat), color = viridis(100), scale='row')
p2 = dittoSeq::dittoDimPlot(spe, var = "major_celltype", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE, legend.show=FALSE) +
    ggtitle("Phenograph clusters on UMAP_mnnCorrected") + guides(fill=guide_legend(ncol=5))

grid.arrange(arrangeGrob(p1$gtable), arrangeGrob(p2), ncol=2)
```


```{r}
props=as.data.frame(table(spe$major_celltype))%>%
    mutate(Freq=Freq/sum(Freq))
ggplot(props, aes(x=Var1, y=Freq, fill=Var1))+
    geom_bar(stat='identity')+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),legend.position = "none") +
    labs(x='Cell type',y='Proportion', title='Cell types proportions after XGBoost classification') 
```


# Save data

```{r}
saveRDS(spe, '../data/non_denoised/spe_cell_annotation_exprs.Rds')
```

## Session info

```{r}
sessionInfo()
```



