---
title: "Spillover correction"
author: "Jacopo Umberto Verga"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
invisible(suppressMessages(c(library(imcRtools),
                 library(tidyverse),
                 library(patchwork),
                 library(scater),
                 library(BiocParallel),
                 library(CATALYST),
                 library(pheatmap),
                 library(cytomapper))))

multicore=MulticoreParam(workers=16)
```


# Import spatial dataset

```{r load_spe}
spe <- readRDS("/home/rstudio/workspace/data/emd/spe.rds")
spe
```




## Single-cell data compensation

The `CATALYST` package can be used to perform spillover compensation on the
**single-cell mean intensities**. Here, the `SpatialExperiment` object generated
in in the previous Session is read in. The `CATALYST` package requires an entry
to `rowData(spe)$channel_name` for the `compCytof` function to run. This entry
should contain the metal isotopes in the form (mt)(mass)Di (e.g., `Sm152Di` for
Samarium isotope with the atomic mass 152). 

The `compCytof` function performs channel spillover compensation on the mean
pixel intensities per channel and cell. Here, we will not overwrite the assays
in the `SpatialExperiment` object to later highlight the effect of compensation.
As shown in the previous session, also the compensated counts are
asinh-transformed using a cofactor of 1.

```{r single-cell-compensation}

sm <- read.csv('/home/rstudio/workspace/data/sm.csv', header = TRUE,  check.names = FALSE, row.names = '')
sm <- sm[rownames(sm)]
sm[is.na(sm)] <- 0
sm <- sm/100
sm <- as.matrix(sm)
diag(sm) <- 1
# TODO drop not used columns
channel_names <- paste0(gsub("[()]","", rowData(spe)$channel),"Di")
rowData(spe)$channel_name <- channel_names
# rownames(sm) <- channel_names
# colnames(sm) <- channel_names

isotope_list <- CATALYST::isotope_list
isotope_list$Ar <- 80


spe <- compCytof(spe, sm, 
                 transform = TRUE, cofactor = 1,
                 isotope_list = isotope_list, 
                 overwrite = FALSE)
```     



To check the effect of channel spillover compensation, the expression of markers
that are affected by spillover (e.g., GZMK in channel Gd(158) and aSMA in
channel Nd(142)) can be visualized in form of scatter plots before and after
compensation.

```{r visualize-single-cell-spillover, message=FALSE}
library(dittoSeq)
library(patchwork)

spill1='aSMA'
spill2='GranzymeK'

before <- dittoScatterPlot(spe, x.var = spill1, y.var = spill2,
                           assay.x = "exprs", assay.y = "exprs") +
    ggtitle("Before compensation")

after <- dittoScatterPlot(spe, x.var = spill1, y.var = spill2,
                          assay.x = "compexprs", assay.y = "compexprs") +
    ggtitle("After compensation")
before + after
```

We observe that the spillover Nd(142) --> Gd(158) was successfully corrected.
To facilitate further downstream analysis, the non-compensated assays can now be
replaced by their compensated counterparts:

```{r overwrite-assays}
assay(spe, "counts") <- assay(spe, "compcounts") 
assay(spe, "exprs") <- assay(spe, "compexprs") 
assay(spe, "compcounts") <- assay(spe, "compexprs") <- NULL
```

```{r save-objects-read-data}
saveRDS(spe, "/home/rstudio/workspace/data/emd/spe_spllover_corrected.rds")
```
